An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2019-11-19 manuscript/chapter_1.tex
Part way through editing

A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}. Ecosystem functions can be defined in broad terms as rate processes and properties which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance \citep{}. Ecosystem functionality may also include static ecosystem properties such as forest canopy complexity and trophic complexity, which in turn influence ecosystem processes. The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support previous findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function has been attributed to niche complementarity \citep{}, whereby competition between individuals is reduced due to them occupying different niche space, allowing diverse species assemblages to take greater advantage of natural resources \citep{}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as much of a role in the relationship between biodiversity and ecosystem function. Instead, stress tolerance and productivity of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{}.

\todo{Re-assess this paragraph -}In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity. This has in turn raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}. Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}. Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016}. The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}. These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, and, others}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends could not be reliably inferred. \todo{Say how the previous studies might have been wrong because of their limited scope, or how in order to really answer the question you need to sample long environmental gradients.}

The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}. In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales in savannas. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum. However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.

\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions. In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}. Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.

Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic rates. Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}. In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas. The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it. Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition. Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness \textit{per se}.

@@

\todo{Species abundance evenness and selection effects.}

\todo{Stand structural complexity}

While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}. \citet{Solbrig1996} however, writes that they are structurally diverse in the tree layer, with trees of the same species occupying distinct layers of the canopy at different growth stages. This structural diversity may be one mechanism through which species diversity appears to influence woody biomass. Trees occupying different canopy layers can create a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area. An underpinning principle of the Biodiversity-Ecosystem Function Relationship (BEFR) is that of niche complementarity. Species are assumed to differ in their functional niche by the fact that they may coexist in a steady-state ecological community, therefore the more species are present the greater the area of the total fundamental niche of the ecosystem is filled, allowing more efficient use of natural resources and higher gross primary productivity and biomass \citep{}. This theory however, which has been supported by many experiments and observational studies in temperate and wet tropical ecosystems, may not hold in savannas, which are structured by disturbance rather than competition.

@@

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{}). We investigated the relationship between aboveground woody biomass and tree biodiversity. We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature. We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type clusters based on tree species composition \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality. Initially, we made four hypotheses:

@@

\begin{enumerate}
	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.} 
	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species diversity.}
	\item{Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.}
	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
\end{enumerate}

@@

The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}). The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

@@

Climatic data was collected from the ECMWF ERA5 dataset \citep{}. Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of the daily data across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil. These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

Observed variables were standardized to Z-scores for the SEMs. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed. Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.6\textwidth]{corr_mat}
	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation. Variables in bold on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable. Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero. Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity. See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
	\label{corr_mat}
\end{figure}

@@

In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density, but not via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}). Stand structural diversity had a positive direct effect on AGB independent of species diversity. Tree species diversity had a strong positive effect on stem density. Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes}
	\caption{Standardized path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Due to all observed variables being standardized and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{struc_model_slopes}
\end{figure}

@@

When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardized path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients. It should be noted that models for vegetation clusters 4 and 5 did not converge, possibly due to small sample sizes (\autoref{struc_model_slopes_all}).

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes_all}
	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Path coefficients are +/- 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{struc_model_slopes_all}
\end{figure}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{full_model_slopes}
	\caption{Standardized path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots. Path coefficients are +/- 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{full_model_slopes}
\end{figure}

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, and [d] stem density on aboveground woody biomass (AGB) across southern African woodlands. Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB, and a relationship between tree structural diversity and AGB that was independent of tree species diversity. Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB.

@@

The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in environment+diversity SEM model (\autoref{full_mod}). Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{}.

Most of the total species diversity effect was through the indirect effect via stem density. This suggests that a higher species diversity allows for a greater density of tree stems and that in southern African woodlands, this leads to an increase in total AGB. The effect of increased stem density within forests is variable. The question of whether lots of small stems or a few large stems maximises AGB has yet to be solved \citep{}.

While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{}. Mechanistically, a higher standing woody biomass results in a greater stand level leaf area index (LAI) \citep{}, with leaf area being a primary determinant of productivity \citep{}.

As tree species diversity increased, this allowed for greater stem density which in turn led to greater AGB.

@@

We found no evidence that species diversity led to an increase in structural diversity. Structural diversity did however have a positive effect on AGB. Our original hpyothesis was that an increase in tree species richness would increase the number of functional forms of trees per unit area due to niche complementarity in growth form of coexisting species. However, it seems more likely that structural diversity is instead merely a result of disturbance history. With highly disturbed plots generally having a less structurally diverse canopy. In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges. In highly disturbed woodland plots, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained.

>>>>

A large number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Liang2016, Cardinale2009}. The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance, and even to static properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.

The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}. It is probable however, that the strength and relative importance of the BEFR varies with ecosystem studied and environmental conditions, with many ecosystems remaining understudied in the BEFR literature.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage, being an oceanic island with many endemic species. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, MORE}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function. Instead, stress tolerance and the functional contribution of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.

Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}). The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{}. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}. In these seasonally dry woodlands, human actions are driving rapid changes in biodiversity. Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}. This has raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}. Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{ANOTHER, Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{}. While rapid biodiversity change is being observed in southern African woodlands, research into the relationship between biodiversity and ecosystem functionality remains scarce.

A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass. Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{}. Small scale studies may therefore be over-estimating the effect of diversity.

%These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.

In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{}, so it is important to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on biomass in analyses at large spatial scales in woodlands. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum. However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.

Temperature and water availability impose physiological limits on woody growth rate by limiting metabolic rates. Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}. In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas. The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it with woodlands existing at the climatic extremes of their bioclimatic range experiencing greater reductions in productivity and standing biomass \citep{}. Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition. Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness.

Environmental conditions may indirectly moderate the observed relationship between biodiversity and ecosystem function. \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in Panamanian tropical forests. Dry season length may therefore indirectly affect ecosystem level productivity in forest ecosystems and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions. In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggested variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this moderation relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}. Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.

@@

While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}. \citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and between species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity. This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{}, may not occur in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nulligying the effects of species diversity on structural diversity.

@@

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{plot_loc}, \autoref{temp_precip_hull}). We investigated the relationship between aboveground woody biomass and tree biodiversity. We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation: precipitation, temperature and soil fertility. We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type clusters based on tree species composition \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality. Initially, we made four hypotheses:

@@

\begin{enumerate}
	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species diversity on above-ground woody biomass.} 
	\item{There will be a positive effect of precipitation and soil fertility on AGB. Part of this positive effect will exist as a mediating effect through tree species diversity.}
	\item{Tree structural diversity of the woodland will interact with tree species diversity to provide an indirect path of influence between species diversity and woody biomass stocks.}
	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
\end{enumerate}

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{SEOSAW} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}). The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

@@

Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of the daily data across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil. These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

Observed variables were standardised to Z-scores for the SEMs. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed. Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.6\textwidth]{corr_mat}
	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation. Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable. Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero. Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity. See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
	\label{corr_mat}
\end{figure}

@@

In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density, but not via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}). Stand structural diversity had a positive direct effect on AGB independent of species diversity. Tree species diversity had a strong positive effect on stem density. Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}). The R-squared of AGB was \strucrsq{}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes}
	\caption{Standardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Due to all observed variables being standardised and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{struc_model_slopes}
\end{figure}

@@

When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients. All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.

Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} had no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).

The strongest total effect of tree species diversity on AGB was in cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.

The greatest effect of structural diversity was found in cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}. Cluster 4 also had the highest median above-ground biomass and the highest median stem density (Table \autoref{clust_summ}). The effects of structural diversity on AGB were highly variable.

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes_all}
	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Path coefficients are $\pm$ 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{struc_model_slopes_all}
\end{figure}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{full_model_slopes}
	\caption{Standardised path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots. Path coefficients are $\pm$ 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{full_model_slopes}
\end{figure}

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, and [d] stem density on aboveground woody biomass (AGB) across southern African woodlands. Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB, and a relationship between tree structural diversity and AGB that was independent of tree species diversity. The strongest effect on AGB was that of stem density, which itself was highly dependent on species diversity. Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB. We did find however that there were weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB. We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB.

@@

The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}). Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.

Most of the total species diversity effect was through the indirect effect via stem density. This suggests that within southern African woodlands a higher species diversity allows for a greater density of tree stems and this leads to an increase in total AGB. While reverse causation with increased stem density causing higher species richness through an increased probability of encountering new species, we accounted for this variation in the extrapolation of species richness according to stem density (sampling effort), suggesting that the observed direction of causation is real. We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}. We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level.

% The question of whether lots of small stems or a few large stems maximises AGB has yet to be solved \citep{}.

While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{}. This suggests that as has been found in many other woodland/forest ecosystems, woody biomass in southern African woodlands can be maximised by increasing species diversity.

@@

We found no evidence that species diversity led to an increase in structural diversity. Structural diversity did however have a positive effect on AGB. Our original hypothesis was that an increase in tree species richness would increase the diversity of available tree functional forms within a plot due and that this would act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light. Given our results however, it seems more likely that structural diversity is instead merely a result of disturbance history, with highly disturbed plots generally having a less structurally diverse canopy \citep{}. In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges. In highly disturbed woodland plots, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained \citep{}. Unfortunately we did not have access to a dataset of high enough spatial and temporal resolution to adequately quantify the effects of disturbance history.
